---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
summary(gapminder)
glimpse(gapminder)
```

**Observations**:

- country, continent, year, lifeExp (life expectancy), pop (population), gdpPercap (gdp per capita)

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(3) %>% max()
year_min <- gapminder %>% pull(3) %>% min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder%>% filter(year==year_min)%>%
  filter_out(gdpPercap>90000)%>%
ggplot(aes(continent,gdpPercap))+
geom_boxplot()
```

**Observations**:

- Write your observations here

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them
Theres an outlier in Asia significantly higher than any other data points which makes all the other points look squished.
To make a readable graph I filtered out points over 90,000.


### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2


cont<- c("Asia", "Africa", 'Europe','Americas', 'Oceania')
for (x in conts) {
 contGdp<-dfYrMin%>%filter(continent==x,year==year_min)
 out<-boxplot.stats(contGdp$gdpPercap)$out
 outInd <- which(contGdp$gdpPercap %in% c(out))
 print(contGdp[outInd,c(1,6)])
 
}

```

**Observations**:

- Identify the outlier countries from q2
  - Bahrain, Kuwait, Angola, Gabon, South Africa, Switzerland, Canada, United States, Venezuela

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder%>% filter(year==c(year_min,year_max))%>%
ggplot(aes(continent,gdpPercap))+
geom_boxplot()+
  facet_grid('year')+
  scale_y_continuous(trans = 'log10')+
    geom_point(
    data = . %>% filter(country %in% c('Bahrain', 'Kuwait', 'Angola', 'Gabon', 'South Africa', 'Switzerland', 'Canada', 'United States', 'Venezuela')),
    mapping = aes(color = country),
    size = 2
  )

```

**Observations**:

- There was only one new outlier in the 2007 graph, compared to the 1952 graph. Kuwait which was an extreme outlier in 1952 was much closer to the averages in 2007. 

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder%>% filter(year==c(year_min,year_max))%>%
ggplot(aes(continent,lifeExp))+
geom_boxplot()+
  facet_grid('year')
cont<- c("Asia", "Africa", 'Europe','Americas', 'Oceania')
for (x in conts) {
 dfCont<-gapminder%>%filter(continent==x,year==year_min)
 out<-boxplot.stats(dfCont$lifeExp)$out
 outInd <- which(dfCont$lifeExp %in% c(out))
 print(dfCont[outInd,c(1,4)])
 
}

for (x in conts) {
 dfCont<-gapminder%>%filter(continent==x,year==year_min)
 out<-boxplot.stats(dfCont$lifeExp)$out
 outInd <- which(dfCont$lifeExp %in% c(out))
 print(dfCont[outInd,c(1,4)])
 
}
  
```

- (Your notes and observations here)
Life expectancies for all continents have increased over time. Africa has a significantly lower life expectancy than any other continent in 2007, but was much closer to Asia and the Americas in 1952
```{r q5-task2}
## TASK: Your second graph
gapminder%>% filter(year>=2000)%>%
ggplot(aes(gdpPercap,lifeExp))+
  geom_point()
```

q: Is life expectancy related to a country's GDP per capita?
note: I decided to use data from 2000 onwards, because I know that both GDP per capita and life expectancy have been increasing over time 
(Your notes and observations here)
Life expectancy increases with a countries GDP although the rate of growth drops off sharply for a gdp higher than around 5000.

```{r q5-task3}
## TASK: Your third graph
gapminder%>% filter(year>=2000)%>%
ggplot(aes(pop,lifeExp))+
  geom_point()+
  scale_x_continuous(trans = 'log10')

```
```

- q: does total population affect the life expectancy of a country
(Your notes and observations here)
Used log scale for x axis as a few outliers made the rest of the data overlap. There is no visible relationship between population and life expectancy. 


